---
title: "In-class Exercise 07: Visualizing, Analyzing, and Forecasting Time Series Data: tidyverts methods"
format: html
author: "Sei Sar Hla Kyi"
date: "`r Sys.Date()`"
date-modified: "`r Sys.Date()`"
execute: 
  eval: true
  echo: true
  message: false
  freeze: true
  warning: false
---

# 1 Install and Load the R Packages

```{r}
pacman::p_load(readr,tidyverse, tsibble, feasts, fable, seasonal)
```

# 2 Data Preparation

## 2.1 Importing the Data

```{r}
ts_data <- read_csv(
  "data/visitor_arrivals_by_air.csv")
```

`dmy()` of **lubridate** package is used to convert data type of Month-Year field from Character to Date.

```{r}
ts_data$`Month-Year` <- dmy(
  ts_data$`Month-Year`)
```

## **2.2 Conventional base `ts` object versus `tibble` object**

Convert to ts object after cleaning the data.

```{r}
ts_data_ts <- ts(ts_data)     
class (ts_data_ts)
```

## **2.3 Converting `tibble` object to `tsibble` object**

```{r}
ts_tsibble <- ts_data %>%
  mutate(Month = yearmonth(`Month-Year`)) %>%
  as_tsibble(index = `Month`)
class (ts_tsibble)
```

# **3 Visualising Time-series Data**

We need to organise the data frame from wide to long format by using [`pivot_longer()`](https://tidyr.tidyverse.org/reference/pivot_longer.html) of **tidyr** package as shown below.

```{r}
ts_longer <- ts_data %>%
  pivot_longer(cols = c(2:34),
               names_to = "Country",
               values_to = "Arrivals")
```

## **3.1 Visualising single time-series: ggplot2 methods**

```{r}
ts_longer %>%
  filter(Country == "Vietnam") %>%
  ggplot(aes(x = `Month-Year`, 
             y = Arrivals))+
  geom_line(size = 0.5)
```

## **3.2 Plotting multiple time-series data with ggplot2 methods**

```{r}
#| fig-height: 8
ggplot(data = ts_longer, 
       aes(x = `Month-Year`, 
           y = Arrivals,
           color = Country))+
  geom_line(size = 0.5) +
  theme(legend.position = "bottom", 
        legend.box.spacing = unit(0.5, "cm"))
```

```{r}
#| fig-height: 12
ggplot(data = ts_longer, 
       aes(x = `Month-Year`, 
           y = Arrivals))+
  geom_line(size = 0.5) +
  facet_wrap(~ Country,
             ncol = 3,
             scales = "free_y") +
  theme_bw()
```

# **4 Visual Analysis of Time-series Data**

```{r}
tsibble_longer <- ts_tsibble %>%
  pivot_longer(cols = c(2:34),
               names_to = "Country",
               values_to = "Arrivals")
```

## **4.1 Visual Analysis of Seasonality with Seasonal Plot**

```{r}
#| fig-height: 12
tsibble_longer %>%
  filter(Country == "Italy" |
         Country == "Vietnam" |
         Country == "United Kingdom" |
         Country == "Germany") %>% 
  gg_season(Arrivals)
```

## **4.2 Visual Analysis of Seasonality with Cycle Plot**

```{r}
#| fig-height: 6
tsibble_longer %>%
  filter(Country == "Vietnam" |
         Country == "Italy") %>% 
  gg_subseries(Arrivals)
```

Visitors from Vietnam has an increasing trend. July and August is quite high. For Italy, for most of the months it is at a low level (hover around 5000). Only in the month of August, there is a sharp increase.

# **5 Time series Analysis - decomposition**

## **5.1 Single time series decomposition**

Autocorrelation Plot: Whether the time series is related to previous month. Correlation among time series(T) and their lags (T-1)

```{r}
tsibble_longer %>%
  filter(`Country` == "Vietnam") %>%
  ACF(Arrivals) %>% 
  autoplot()
```

## **5.2 Multiple time-series decomposition**

```{r}
tsibble_longer %>%
  filter(`Country` == "Vietnam" |
         `Country` == "Italy" |
         `Country` == "United Kingdom" |
         `Country` == "China") %>%
  ACF(Arrivals) %>%
  autoplot()
```

Autocorrelation of China is in a 6 month period.

Look at 95% confidence interval (above the blue line): For China and Vietnam, most of the cases are statistically significant.

There are both seasonality and trend in China and Vietnam.

```{r}
tsibble_longer %>%
  filter(`Country` == "Vietnam" |
         `Country` == "Italy" |
         `Country` == "United Kingdom" |
         `Country` == "China") %>%
  PACF(Arrivals) %>%
  autoplot()
```

Partial Autocorrelation: Further decompose the autocorrelation, Compare autocorrelation from previous time lag versus next time lag. Inverse(negative): visitor arrival decrease.

For UK, in lag 2 significantly reduced.

# **6 Visual STL Diagnostics**

## **6.1 Visual STL diagnostics with feasts**

```{r}
tsibble_longer %>%
  filter(`Country` == "Vietnam") %>%
  model(stl = STL(Arrivals)) %>%
  components() %>%
  autoplot()
```

Remainder: white noise. No clear pattern

If Remainder still give patterns, it is not good.

# **7 Visual Forecasting**

## **7.1 Time Series Data Sampling**

Intentionally keep back some data based on time, must follow the sequence.

2 Levels: Holdout and Training

```{r}
vietnam_ts <- tsibble_longer %>%
  filter(Country == "Vietnam") %>% 
  mutate(Type = if_else(
    `Month-Year` >= "2019-01-01", 
    "Hold-out", "Training"))
```

```{r}
vietnam_train <- vietnam_ts %>%
  filter(`Month-Year` < "2019-01-01")
```

## **7.2 Fitting forecasting models**

```{r}
fit_ses <- vietnam_train %>%
  model(ETS(Arrivals ~ error("A") 
            + trend("N") 
            + season("N")))
gg_tsresiduals(fit_ses)
```

```{r}

```

```{r}

```

```{r}
fit_ETS <- vietnam_train %>%
  model(`SES` = ETS(Arrivals ~ error("A") + 
                      trend("N") + 
                      season("N")),
        `Holt`= ETS(Arrivals ~ error("A") +
                      trend("A") +
                      season("N")),
        `damped Holt` = 
          ETS(Arrivals ~ error("A") +
                trend("Ad") + 
                season("N")),
        `WH_A` = ETS(
          Arrivals ~ error("A") + 
            trend("A") + 
            season("A")),
        `WH_M` = ETS(Arrivals ~ error("M") 
                         + trend("A") 
                         + season("M"))
  )
```

```{r}
fit_ETS %>%
  tidy()
```

## 7.3 Model Comparison

```{r}
fit_ETS %>% 
  report()
```

## **7.4 Forecasting future values**

```{r}
fit_ETS %>%
  forecast(h = "12 months") %>%
  autoplot(vietnam_ts, 
           level = NULL)
```
