---
title: "Hands-on Exercise 03: Programming Interactive Data Visualisation and Animated Statistical Graphics with R"
format: html
author: "Sei Sar Hla Kyi"
date: "`r Sys.Date()`"
date-modified: "`r Sys.Date()`"
execute: 
  eval: true
  echo: true
  message: false
  freeze: true
  warning: false
---

# 1 Programming Interactive Data Visualisation with R 

## 1.1 Installing and Loading Packages

::: panel-tabset
## Explanation of Packages

The following packages will be installed for this exercise:

-   [**ggiraph**](https://davidgohel.github.io/ggiraph/) for making ‘ggplot’ graphics interactive.

-   [**plotly**](https://plotly.com/r/), R library for plotting interactive statistical graphs.

-   [**DT**](https://rstudio.github.io/DT/) provides an R interface to the JavaScript library [DataTables](https://datatables.net/) that create interactive table on html page.

-   [**tidyverse**](https://www.tidyverse.org/), a family of modern R packages specially designed to support data science, analysis and communication task including creating static statistical graphs.

-   [**patchwork**](https://patchwork.data-imaginist.com/) for combining multiple ggplot2 graphs into one figure.

## Code

```{r}
pacman::p_load(ggiraph, plotly, 
               patchwork, DT, tidyverse) 
```
:::

## 1.2 Importing Data

The code chunk below `read_csv()` of **readr** package is used to import *Exam_data.csv* data file into R and save it as an tibble data frame called `exam_data`.

```{r}
exam_data <- read_csv("data/Exam_data.csv")
```

## 1.3 Interactive Data Visualisation - ggiraph methods

[ggiraph](https://davidgohel.github.io/ggiraph/index.html) is an htmlwidget and a ggplot2 extension. It allows ggplot graphics to be interactive.

Interactive is made with [**ggplot geometries**](https://davidgohel.github.io/ggiraph/reference/index.html#section-interactive-geometries) that can understand three arguments:

-   **Tooltip**: a column of data-sets that contain tooltips to be displayed when the mouse is over elements.

-   **Onclick**: a column of data-sets that contain a JavaScript function to be executed when elements are clicked.

-   **Data_id**: a column of data-sets that contain an id to be associated with elements.

::: panel-tabset
## Explanation of Code

The code chunk consists of two parts. First, an ggplot object will be created. Next, [`girafe()`](https://davidgohel.github.io/ggiraph/reference/girafe.html) of **ggiraph** will be used to create an interactive svg object.

Two steps are involved. First, an interactive version of ggplot2 geom (i.e. [`geom_dotplot_interactive()`](https://davidgohel.github.io/ggiraph/reference/geom_dotplot_interactive.html)) will be used to create the basic graph. Then, [`girafe()`](https://davidgohel.github.io/ggiraph/reference/girafe.html) will be used to generate an svg object to be displayed on an html page.

## Code

```{r}
p <- ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_dotplot_interactive(
    aes(tooltip = ID),
    stackgroups = TRUE, 
    binwidth = 1, 
    method = "histodot") +
  scale_y_continuous(NULL, 
                     breaks = NULL)
girafe(
  ggobj = p,
  width_svg = 6,
  height_svg = 6*0.618
)
```
:::

### 1.3.1 Displaying multiple information on tooltip

The content of the tooltip can be customised by including a list object as shown in the code chunk below.

::: callout-note
The first three lines of codes in the code chunk create a new field called *tooltip*. At the same time, it populates text in ID and CLASS fields into the newly created field. Next, this newly created field is used as tooltip field as shown in the code of line 7.
:::

```{r}
exam_data$tooltip <- c(paste0(     
  "Name = ", exam_data$ID,         
  "\n Class = ", exam_data$CLASS)) 

p <- ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_dotplot_interactive(
    aes(tooltip = exam_data$tooltip), 
    stackgroups = TRUE,
    binwidth = 1,
    method = "histodot") +
  scale_y_continuous(NULL,               
                     breaks = NULL)
girafe(
  ggobj = p,
  width_svg = 8,
  height_svg = 8*0.618
)
```

### 1.3.2 Customising Tooltip style

Code chunk below uses [`opts_tooltip()`](https://davidgohel.github.io/ggiraph/reference/opts_tooltip.html) of **ggiraph** to customize tooltip rendering by add css declarations.

```{r}
tooltip_css <- "background-color:white; #<<
font-style:bold; color:black;" #<<

p <- ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_dotplot_interactive(              
    aes(tooltip = ID),                   
    stackgroups = TRUE,                  
    binwidth = 1,                        
    method = "histodot") +               
  scale_y_continuous(NULL,               
                     breaks = NULL)
girafe(                                  
  ggobj = p,                             
  width_svg = 6,                         
  height_svg = 6*0.618,
  options = list(    #<<
    opts_tooltip(    #<<
      css = tooltip_css)) #<<
)                                        
```

### 1.3.3 Displaying statistics on tooltip

Code chunk below shows an advanced way to customise tooltip. In this example, a function is used to compute 90% confident interval of the mean. The derived statistics are then displayed in the tooltip.

```{r}
tooltip <- function(y, ymax, accuracy = .01) {
  mean <- scales::number(y, accuracy = accuracy)
  sem <- scales::number(ymax - y, accuracy = accuracy)
  paste("Mean maths scores:", mean, "+/-", sem)
}

gg_point <- ggplot(data=exam_data, 
                   aes(x = RACE),
) +
  stat_summary(aes(y = MATHS, 
                   tooltip = after_stat(  
                     tooltip(y, ymax))),  
    fun.data = "mean_se", 
    geom = GeomInteractiveCol,  
    fill = "light blue"
  ) +
  stat_summary(aes(y = MATHS),
    fun.data = mean_se,
    geom = "errorbar", width = 0.2, size = 0.2
  )

girafe(ggobj = gg_point,
       width_svg = 8,
       height_svg = 8*0.618)
```

### 1.3.4 Hover effect with *data_id* aesthetic

```{r}
p <- ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_dotplot_interactive(           
    aes(data_id = CLASS),             
    stackgroups = TRUE,               
    binwidth = 1,                        
    method = "histodot") +               
  scale_y_continuous(NULL,               
                     breaks = NULL)
girafe(                                  
  ggobj = p,                             
  width_svg = 6,                         
  height_svg = 6*0.618                      
)                                        
```

::: callout-note
Interactivity: Elements associated with a *data_id* (i.e CLASS) will be highlighted upon mouse over.default value of the hover css is *hover_css = “fill:orange;”*.
:::

### 1.3.5 Styling hover effect

```{r}
p <- ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_dotplot_interactive(              
    aes(data_id = CLASS),              
    stackgroups = TRUE,                  
    binwidth = 1,                        
    method = "histodot") +               
  scale_y_continuous(NULL,               
                     breaks = NULL)
girafe(                                  
  ggobj = p,                             
  width_svg = 6,                         
  height_svg = 6*0.618,
  options = list(                        
    opts_hover(css = "fill: #202020;"),  
    opts_hover_inv(css = "opacity:0.2;") 
  )                                        
)                                        
```

::: callout-note
Interactivity: Elements associated with a *data_id* (i.e CLASS) will be highlighted upon mouse over.Different from previous example, in this example the ccs customisation request are encoded directly.
:::

### 1.3.6 Combining tooltip and hover effect

```{r}
p <- ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_dotplot_interactive(              
    aes(tooltip = CLASS, 
        data_id = CLASS),              
    stackgroups = TRUE,                  
    binwidth = 1,                        
    method = "histodot") +               
  scale_y_continuous(NULL,               
                     breaks = NULL)
girafe(                                  
  ggobj = p,                             
  width_svg = 6,                         
  height_svg = 6*0.618,
  options = list(                        
    opts_hover(css = "fill: #202020;"),  
    opts_hover_inv(css = "opacity:0.2;") 
  )                                        
)                                        
```

::: callout-note
Interactivity: Elements associated with a *data_id* (i.e CLASS) will be highlighted upon mouse over. At the same time, the tooltip will show the CLASS.
:::

### 1.3.7 Click effect with onclick

`onclick` argument of ggiraph provides hotlink interactivity on the web.

```{r}
exam_data$onclick <- sprintf("window.open(\"%s%s\")",
"https://www.moe.gov.sg/schoolfinder?journey=Primary%20school",
as.character(exam_data$ID))

p <- ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_dotplot_interactive(              
    aes(onclick = onclick),              
    stackgroups = TRUE,                  
    binwidth = 1,                        
    method = "histodot") +               
  scale_y_continuous(NULL,               
                     breaks = NULL)
girafe(                                  
  ggobj = p,                             
  width_svg = 6,                         
  height_svg = 6*0.618)                                        
```

::: callout-note
Interactivity: Web document link with a data object will be displayed on the web browser upon mouse click.
:::

### 1.3.8 Coordinated Multiple Views with ggiraph

Coordinated multiple views - when a data point of one of the dotplot is selected, the corresponding data point ID on the second data visualisation will be highlighted too.

In order to build a coordinated multiple views, the following programming strategy will be used:

1.  Appropriate interactive functions of **ggiraph** will be used to create the multiple views.

2.  *patchwork* function of [patchwork](https://patchwork.data-imaginist.com/) package will be used inside girafe function to create the interactive coordinated multiple views.

```{r}
p1 <- ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_dotplot_interactive(              
    aes(data_id = ID),              
    stackgroups = TRUE,                  
    binwidth = 1,                        
    method = "histodot") +  
  coord_cartesian(xlim=c(0,100)) + 
  scale_y_continuous(NULL,               
                     breaks = NULL)

p2 <- ggplot(data=exam_data, 
       aes(x = ENGLISH)) +
  geom_dotplot_interactive(              
    aes(data_id = ID),              
    stackgroups = TRUE,                  
    binwidth = 1,                        
    method = "histodot") + 
  coord_cartesian(xlim=c(0,100)) + 
  scale_y_continuous(NULL,               
                     breaks = NULL)

girafe(code = print(p1 + p2), 
       width_svg = 6,
       height_svg = 3,
       options = list(
         opts_hover(css = "fill: #202020;"),
         opts_hover_inv(css = "opacity:0.2;")
         )
       ) 
```

## 1.4 Interactive Data Visualisation - plotly methods

There are two ways to create interactive graph by using plotly, they are:

-   by using *plot_ly()*, and

-   by using *ggplotly()*

### 1.4.1 Creating an interactive scatter plot: plot_ly() method

The code chunk below shows a basic interactive scatter plot made from plot_ly

```{r}
plot_ly(data = exam_data, 
             x = ~MATHS, 
             y = ~ENGLISH)
```

### 1.4.2 Working with visual variable: plot_ly() method

*Color* argument is mapped to a qualitative visual variable (i.e. RACE).

```{r}
plot_ly(data = exam_data, 
        x = ~ENGLISH, 
        y = ~MATHS, 
        color = ~RACE)
```

### 1.4.3 Creating an interactive scatter plot: ggplotly() method

```{r}
p <- ggplot(data=exam_data, 
            aes(x = MATHS,
                y = ENGLISH)) +
  geom_point(size=1) +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100))
ggplotly(p)
```

::: callout-note
Notice that the only extra line you need to include in the code chunk is *ggplotly()*.
:::

### 1.4.4 Coordinated Multiple Views with plotly

::: panel-tabset
## Explanation of Code

The creation of a coordinated linked plot by using plotly involves three steps:

-   [`highlight_key()`](https://www.rdocumentation.org/packages/plotly/versions/4.9.2/topics/highlight_key) of **plotly** package is used as shared data. This simply creates an object of class [crosstalk::SharedData](https://rdrr.io/cran/crosstalk/man/SharedData.html)

-   two scatterplots will be created by using ggplot2 functions.

-   lastly, [*subplot()*](https://plotly.com/r/subplots/) of **plotly** package is used to place them next to each other side-by-side.

## Code

```{r}
d <- highlight_key(exam_data)
p1 <- ggplot(data=d, 
            aes(x = MATHS,
                y = ENGLISH)) +
  geom_point(size=1) +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100))

p2 <- ggplot(data=d, 
            aes(x = MATHS,
                y = SCIENCE)) +
  geom_point(size=1) +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100))
subplot(ggplotly(p1),
        ggplotly(p2))
```
:::

## 1.5 Interactive Data Visualisation - crosstalk methods

[Crosstalk](https://rstudio.github.io/crosstalk/index.html) is an add-on to the htmlwidgets package. It extends htmlwidgets with a set of classes, functions, and conventions for implementing cross-widget interactions (currently, linked brushing and filtering).

### 1.5.1 Interactive Data Table: DT package

::: panel-tabset
## Explanation

-   A wrapper of the JavaScript Library [DataTables](https://datatables.net/)

-   Data objects in R can be rendered as HTML tables using the JavaScript library ‘DataTables’ (typically via R Markdown or Shiny).

## Code

```{r}
DT::datatable(exam_data, class= "compact")
```
:::

### 1.5.2 Linked brushing

-   *highlight()* is a function of **plotly** package. It sets a variety of options for brushing multiple plots. These options are primarily designed for linking multiple plotly graphs, and may not behave as expected when linking plotly to another htmlwidget package via crosstalk. In some cases, other htmlwidgets will respect these options, such as persistent selection in leaflet.

-   *bscols()* is a helper function of **crosstalk** package. It makes it easy to put HTML elements side by side. It can be called directly from the console but is especially designed to work in an R Markdown document. **Warning:** This will bring in all of Bootstrap!.

```{r}
d <- highlight_key(exam_data) 
p <- ggplot(d, 
            aes(ENGLISH, 
                MATHS)) + 
  geom_point(size=1) +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100))

gg <- highlight(ggplotly(p),        
                "plotly_selected")  

crosstalk::bscols(gg,               
                  DT::datatable(d), 
                  widths = 5)        
```

# 2 Programming Animated Statistical Graphics with R 

## 2.1 Basic concepts of animation

When creating animations, the plot does not actually move. Instead, many individual plots are built and then stitched together as movie frames, just like an old-school flip book or cartoon. Each frame is a different plot when conveying motion, which is built using some relevant subset of the aggregate data. The subset drives the flow of the animation when stitched back together.

Some terminology to note:

1.  **Frame**: In an animated line graph, each frame represents a different point in time or a different category. When the frame changes, the data points on the graph are updated to reflect the new data.

2.  **Animation Attributes**: The animation attributes are the settings that control how the animation behaves. For example, you can specify the duration of each frame, the easing function used to transition between frames, and whether to start the animation from the current frame or from the beginning.

::: callout-note
## Ask yourself!

Does it makes sense to go through the effort? If you are conducting an exploratory data analysis, a animated graphic may not be worth the time investment. However, if you are giving a presentation, a few well-placed animated graphics can help an audience connect with your topic remarkably better than static counterparts
:::

## 2.2 Installing and Loading the R packages

::: panel-tabset
## Packages

-   [**plotly**](https://plotly.com/r/), R library for plotting interactive statistical graphs.

-   [**gganimate**](https://gganimate.com/), an ggplot extension for creating animated statistical graphs.

-   [**gifski**](https://cran.r-project.org/web/packages/gifski/index.html) converts video frames to GIF animations using pngquant’s fancy features for efficient cross-frame palettes and temporal dithering. It produces animated GIFs that use thousands of colors per frame.

-   [**gapminder**](https://cran.r-project.org/web/packages/gapminder/index.html): An excerpt of the data available at Gapminder.org. We just want to use its *country_colors* scheme.

-   [**tidyverse**](https://www.tidyverse.org/), a family of modern R packages specially designed to support data science, analysis and communication task including creating static statistical graphs.

## Code

```{r}
pacman::p_load(readxl, gifski, gapminder,
               plotly, gganimate, tidyverse)
```
:::

## 2.3 Importing the data

::: panel-tabset
## Explanation

-   [`read_xls()`](https://readxl.tidyverse.org/reference/read_excel.html) of **readxl** package is used to import the Excel worksheet.

-   `mutate_each_()` of dplyr package is used to convert all character data type into factor. (Since `mutate_each_()` was deprecated in dplyr 0.7.0. and `funs()` was deprecated in dplyr 0.8.0, we will re-write the code by using [`mutate_at()`](https://r4va.netlify.app/chap04) as shown in the code chunk below.

-   `mutate` of dplyr package is used to convert data values of Year field into integer.

## Code

```{r}
col <- c("Country", "Continent")
globalPop <- read_xls("data/GlobalPopulation.xls",
                      sheet="Data") %>%
  mutate_at(col, as.factor) %>%
  mutate(Year = as.integer(Year))
```
:::

##  2.4 Animated Data Visualisation - gganimate methods

[**gganimate**](https://gganimate.com/) extends the grammar of graphics as implemented by ggplot2 to include the description of animation. It does this by providing a range of new grammar classes that can be added to the plot object in order to customise how it should change with time.

-   `transition_*()` defines how the data should be spread out and how it relates to itself across time.

-   `view_*()` defines how the positional scales should change along the animation.

-   `shadow_*()` defines how data from other points in time should be presented in the given point in time.

-   `enter_*()/exit_*()` defines how new data should appear and how old data should disappear during the course of the animation.

-   `ease_aes()` defines how different aesthetics should be eased during transitions.

### 2.4.1 Building a static population bubble plot

```{r}
ggplot(globalPop, aes(x = Old, y = Young, 
                      size = Population, 
                      colour = Country)) +
  geom_point(alpha = 0.7, 
             show.legend = FALSE) +
  scale_colour_manual(values = country_colors) +
  scale_size(range = c(2, 12)) +
  labs(title = 'Year: {frame_time}', 
       x = '% Aged', 
       y = '% Young') 
```

### 2.4.2 Building a animated bubble plot

::: panel-tabset
## Explanation

-   [`transition_time()`](https://gganimate.com/reference/transition_time.html) of **gganimate** is used to create transition through distinct states in time (i.e. Year).

-   `ease_aes()` is used to control easing of aesthetics. The default is `linear`. Other methods are: quadratic, cubic, quartic, quintic, sine, circular, exponential, elastic, back, and bounce.

## Code

```{r}
ggplot(globalPop, aes(x = Old, y = Young, 
                      size = Population, 
                      colour = Country)) +
  geom_point(alpha = 0.7, 
             show.legend = FALSE) +
  scale_colour_manual(values = country_colors) +
  scale_size(range = c(2, 12)) +
  labs(title = 'Year: {frame_time}', 
       x = '% Aged', 
       y = '% Young') +
  transition_time(Year) +       
  ease_aes('linear')          
```
:::

##  2.5 Animated Data Visualisation - plotly methods

In **Plotly R** package, both `ggplotly()` and `plot_ly()` support key frame animations through the `frame` argument/aesthetic. They also support an `ids` argument/aesthetic to ensure smooth transitions between objects with the same id.

### 2.5.1 Building an animated bubble plot: `ggplotly()` method

```{r}
gg <- ggplot(globalPop, 
       aes(x = Old, 
           y = Young, 
           size = Population, 
           colour = Country)) +
  geom_point(aes(size = Population,
                 frame = Year),
             alpha = 0.7, 
             show.legend = FALSE) +
  scale_colour_manual(values = country_colors) +
  scale_size(range = c(2, 12)) +
  labs(x = '% Aged', 
       y = '% Young')

ggplotly(gg)
```

::: callout-note
Although `show.legend = FALSE` argument was used, the legend still appears on the plot. To overcome this problem, `theme(legend.position='none')` should be used as shown in the plot and code chunk below.
:::

```{r}
gg <- ggplot(globalPop, 
       aes(x = Old, 
           y = Young, 
           size = Population, 
           colour = Country)) +
  geom_point(aes(size = Population,
                 frame = Year),
             alpha = 0.7) +
  scale_colour_manual(values = country_colors) +
  scale_size(range = c(2, 12)) +
  labs(x = '% Aged', 
       y = '% Young') + 
  theme(legend.position='none')

ggplotly(gg)
```

### 2.5.2 Building an animated bubble plot: `plotly()` method

```{r}
bp <- globalPop %>%
  plot_ly(x = ~Old, 
          y = ~Young, 
          size = ~Population, 
          color = ~Continent,
          sizes = c(2, 100),
          frame = ~Year, 
          text = ~Country, 
          hoverinfo = "text",
          type = 'scatter',
          mode = 'markers'
          ) %>%
  layout(showlegend = FALSE)
bp
```
