---
title: "Hands-on Exercise 02: Beyond ggplot2 Fundamentals"
format: html
author: "Sei Sar Hla Kyi"
date: "`r Sys.Date()`"
date-modified: "`r Sys.Date()`"
execute: 
  eval: true
  echo: true
  message: false
  freeze: true
  warning: false
---

# **1 Installing and Loading Packages**

::: panel-tabset
## Code

```{r}
pacman::p_load(ggrepel, patchwork, 
               ggthemes, hrbrthemes,
               tidyverse) 
```

## Explanation

In this exercise, beside tidyverse, four R packages will be used. They are:

-   ggrepel: an R package provides geoms for ggplot2 to repel overlapping text labels.

-   ggthemes: an R package provides some extra themes, geoms, and scales for ‘ggplot2’.

-   hrbrthemes: an R package provides typography-centric themes and theme components for ggplot2.

-   patchwork: an R package for preparing composite figure created using ggplot2.
:::

# **2 Importing Data**

The code chunk below imports *exam_data.csv* into R environment by using [*read_csv()*](https://readr.tidyverse.org/reference/read_delim.html) function of [**readr**](https://readr.tidyverse.org/) package. **readr** is one of the tidyverse package.

```{r}
exam_data <- read_csv("data/Exam_data.csv")
```

There are a total of seven attributes in the exam_data tibble data frame.

-   The categorical attributes are: ID, CLASS, GENDER and RACE.

-   The continuous attributes are: MATHS, ENGLISH and SCIENCE.

# **3 Beyond ggplot2 Annotation: ggrepel**

::: panel-tabset
## The Plot

```{r}
#| echo: False
ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              size=0.5) +  
  geom_label(aes(label = ID), 
             hjust = .5, 
             vjust = -.5) +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100)) +
  ggtitle("English scores versus Maths scores for Primary 3")
```

## The Code

```{r}
#| eval: False
ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              size=0.5) +  
  geom_label(aes(label = ID), 
             hjust = .5, 
             vjust = -.5) +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100)) +
  ggtitle("English scores versus Maths scores for Primary 3")
```
:::

[**ggrepel**](https://ggrepel.slowkow.com/) is an extension of **ggplot2** package which provides `geoms` for **ggplot2** to repel overlapping text as in our examples on the right.

::: callout-note
Replace `geom_text()` by [`geom_text_repel()`](https://ggrepel.slowkow.com/reference/geom_text_repel.html) and `geom_label()` by [`geom_label_repel`](https://ggrepel.slowkow.com/reference/geom_text_repel.html).
:::

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              size=0.5) +  
  geom_label_repel(aes(label = ID), 
                   fontface = "bold") +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100)) +
  ggtitle("English scores versus Maths scores for Primary 3")
```

# **4 Beyond ggplot2 Themes**

ggplot2 comes with eight [built-in themes](https://ggplot2.tidyverse.org/reference/ggtheme.html), they are: `theme_gray()`, `theme_bw()`, `theme_classic()`, `theme_dark()`, `theme_light()`, `theme_linedraw()`, `theme_minimal()`, and `theme_void()`

```{r}
ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  theme_dark() +
  ggtitle("Distribution of Maths scores") 
```

## **4.1 Working with ggtheme package**

[**ggthemes**](https://cran.r-project.org/web/packages/ggthemes/) provides [‘ggplot2’ themes](https://yutannihilation.github.io/allYourFigureAreBelongToUs/ggthemes/) that replicate the look of plots by Edward Tufte, Stephen Few, [Fivethirtyeight](https://fivethirtyeight.com/), [The Economist](https://www.economist.com/graphic-detail), ‘Stata’, ‘Excel’, and [The Wall Street Journal](https://www.pinterest.com/wsjgraphics/wsj-graphics/), among others.

```{r}
ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  ggtitle("Distribution of Maths scores") +
  theme_solarized()
```

::: callout-note
The solarized theme is used here.
:::

## **4.2 Working with hrbthems package**

[**hrbrthemes**](https://cinc.rud.is/web/packages/hrbrthemes/) package provides a base theme that focuses on typographic elements, including where various labels are placed as well as the fonts that are used.

```{r}
ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  ggtitle("Distribution of Maths scores") +
  theme_ipsum()
```

Add in the following three:

-   `axis_title_size` argument is used to increase the font size of the axis title to 18,

-   `base_size` argument is used to increase the default axis label to 15, and

-   `grid` argument is used to remove the x-axis grid lines.

```{r}
ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  ggtitle("Distribution of Maths scores") +
  theme_ipsum(axis_title_size = 18,
              base_size = 15,
              grid = "Y")
```

# **5 Beyond Single Graph: patchwork methods**

There are several ggplot2 extensions provide functions to compose figure with multiple graphs.

Let's create three statistical graphics first.

::: panel-tabset
## p1

```{r}
p1 <- ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") + 
  coord_cartesian(xlim=c(0,100)) +
  ggtitle("Distribution of Maths scores")

p1
```

## p2

```{r}
p2 <- ggplot(data=exam_data, 
             aes(x = ENGLISH)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  coord_cartesian(xlim=c(0,100)) +
  ggtitle("Distribution of English scores")

p2
```

## p3

```{r}
p3 <- ggplot(data=exam_data, 
             aes(x= MATHS, 
                 y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              size=0.5) +  
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100)) +
  ggtitle("English scores versus Maths scores for Primary 3")

p3
```
:::

There are several ggplot2 extension’s functions support the needs to prepare composite figure by combining several graphs such as [`grid.arrange()`](https://cran.r-project.org/web/packages/gridExtra/vignettes/arrangeGrob.html) of **gridExtra** package and [`plot_grid()`](https://wilkelab.org/cowplot/reference/plot_grid.html) of [**cowplot**](https://wilkelab.org/cowplot/) package.

There is a ggplot2 extension called [**patchwork**](https://patchwork.data-imaginist.com/) which is specially designed for combining separate ggplot2 graphs into a single figure.

Patchwork package has a very simple syntax where we can create layouts super easily. Here’s the general syntax that combines:

-   Two-Column Layout using the Plus Sign +.

-   Parenthesis () to create a subplot group.

-   Two-Row Layout using the Division Sign `/`

## **5.1 Combining two ggplot2 graphs**

```{r}
p1 + p2
```

## **5.2 Combining three ggplot2 graphs**

-   “/” operator to stack two ggplot2 graphs,

-   “\|” operator to place the plots beside each other,

-   “()” operator the define the sequence of the plotting.

```{r}
(p1 / p2) | p3
```

##  **5.3 Creating a composite figure with tag**

-   `plot_annotation()` is used to add labels, titles, captions, and other metadata to the combined plot.

-   The argument `tag_levels = 'I'` automatically tags the individual subplots with Roman numerals (e.g., I, II, III).

```{r}
((p1 / p2) | p3) + 
  plot_annotation(tag_levels = 'I')
```

## **5.4 Creating figure with insert**

With [`inset_element()`](https://patchwork.data-imaginist.com/reference/inset_element.html) of **patchwork**, we can place one or several plots or graphic elements freely on top or below another plot.

```{r}
p3 + inset_element(p2, 
                   left = 0.02, 
                   bottom = 0.7, 
                   right = 0.5, 
                   top = 1)
```

## 5.5 **Creating a composite figure by using patchwork and ggtheme**

Putting what we learned through this exercise all together:

```{r}
patchwork <- (p1 / p2) | p3
patchwork & theme_solarized()
```
