---
title: "Hands-on Exercise 09:Modelling, Visualising and Analysing Network Data with R"
format: html
author: "Sei Sar Hla Kyi"
date: "`r Sys.Date()`"
date-modified: "`r Sys.Date()`"
execute: 
  eval: true
  echo: true
  message: false
  freeze: true
  warning: false
---

# 1 Installing and launching R packages

```{r}
pacman::p_load(igraph, tidygraph, ggraph, 
               visNetwork, lubridate, clock,
               tidyverse, graphlayouts)
```

# **2 The Data**

**The edges data:** *GAStech-email_edges.csv* which consists of two weeks of 9063 emails correspondances between 55 employees.\
**The nodes data:** *GAStech_email_nodes.csv* which consist of the names, department and title of the 55 employees.

```{r}
GAStech_nodes <- read_csv("data/GAStech_email_node.csv")
GAStech_edges <- read_csv("data/GAStech_email_edge-v2.csv")
```

## **2.1 Reviewing the imported data**

```{r}
glimpse(GAStech_edges)
```

## **2.2 Data Wrangling**

::: panel-tabset
## The Code

```{r}
GAStech_edges <- GAStech_edges %>%
  mutate(SendDate = dmy(SentDate)) %>%
  mutate(Weekday = wday(SentDate,
                        label = TRUE,
                        abbr = FALSE))
```

## Things to note

-   both *dmy()* and *wday()* are functions of **lubridate** package. [lubridate](https://r4va.netlify.app/cran.r-project.org/web/packages/lubridate/vignettes/lubridate.html) is an R package that makes it easier to work with dates and times.

-   *dmy()* transforms the SentDate to Date data type.

-   *wday()* returns the day of the week as a decimal number or an ordered factor if label is TRUE. The argument abbr is FALSE keep the daya spells in full, i.e. Monday. The function will create a new column in the data.frame i.e. Weekday and the output of *wday()* will save in this newly created field.

-   the values in the *Weekday* field are in ordinal scale.
:::

A close examination of *GAStech_edges* data.frame reveals that it consists of individual e-mail flow records. This is not very useful for visualisation.

In view of this, we will aggregate the individual by date, senders, receivers, main subject and day of the week.

::: panel-tabset
## The Code

```{r}
GAStech_edges_aggregated <- GAStech_edges %>%
  filter(MainSubject == "Work related") %>%
  group_by(source, target, Weekday) %>%
    summarise(Weight = n()) %>%
  filter(source!=target) %>%
  filter(Weight > 1) %>%
  ungroup()
```

## Things to note

-   four functions from **dplyr** package are used. They are: *filter()*, *group()*, *summarise()*, and *ungroup()*.

-   The output data.frame is called **GAStech_edges_aggregated**.

-   A new field called *Weight* has been added in GAStech_edges_aggregated
:::

# **3 Creating network objects using tidygraph**

## **3.1 Using `tbl_graph()` to build tidygraph data model**

```{r}
GAStech_graph <- tbl_graph(nodes = GAStech_nodes,
                           edges = GAStech_edges_aggregated, 
                           directed = TRUE)
```

## **3.2 Reviewing the output tidygraph’s graph object**

```{r}
GAStech_graph
```

-   The output above reveals that *GAStech_graph* is a tbl_graph object with 54 nodes and 4541 edges.

-   The command also prints the first six rows of “Node Data” and the first three of “Edge Data”.

-   It states that the Node Data is **active**. The notion of an active tibble within a tbl_graph object makes it possible to manipulate the data in one tibble at a time.

## **3.3 Changing the active object**

```{r}
GAStech_graph %>%
  activate(edges) %>%
  arrange(desc(Weight))
```

If we wanted to rearrange the rows in the edges tibble to list those with the highest “weight” first, we could use *activate()* and then *arrange()*.

# **4 Plotting Static Network Graphs with ggraph package**

## **4.1 Plotting a basic network graph**

```{r}
ggraph(GAStech_graph) +
  geom_edge_link() +
  geom_node_point()
```

## **4.2 Changing the default network graph theme**

```{r}
g <- ggraph(GAStech_graph) + 
  geom_edge_link(aes()) +
  geom_node_point(aes())

g + theme_graph()
```

-   **ggraph** introduces a special ggplot theme that provides better defaults for network graphs than the normal ggplot defaults. `theme_graph()`, besides removing axes, grids, and border, changes the font to Arial Narrow (this can be overridden).

-   The ggraph theme can be set for a series of plots with the `set_graph_style()` command run before the graphs are plotted or by using `theme_graph()` in the individual plots.

## **4.3 Changing the coloring of the plot**

```{r}
g <- ggraph(GAStech_graph) + 
  geom_edge_link(aes(colour = 'grey50')) +
  geom_node_point(aes(colour = 'grey40'))

g + theme_graph(background = 'grey10',
                text_colour = 'white')
```

## **4.4 Working with ggraph’s layouts**

### **4.4.1 Fruchterman and Reingold layout**

```{r}
g <- ggraph(GAStech_graph, 
            layout = "fr") +
  geom_edge_link(aes()) +
  geom_node_point(aes())

g + theme_graph()
```

### **4.4.2 Modifying network nodes**

```{r}
g <- ggraph(GAStech_graph, 
            layout = "nicely") + 
  geom_edge_link(aes()) +
  geom_node_point(aes(colour = Department, 
                      size = 3))

g + theme_graph()
```

-   *geom_node_point* is equivalent in functionality to *geo_point* of **ggplot2**. It allows for simple plotting of nodes in different shapes, colours and sizes. In the codes chnuks above colour and size are used.

### **4.4.3 Modifying edges**

```{r}
g <- ggraph(GAStech_graph, 
            layout = "nicely") +
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department), 
                  size = 3)

g + theme_graph()
```

-   *geom_edge_link* draws edges in the simplest way - as straight lines between the start and end nodes. But, it can do more that that. In the example above, argument *width* is used to map the width of the line in proportional to the Weight attribute and argument alpha is used to introduce opacity on the line.

## **4.5 Creating facet graphs**

### 4.5.1 **Working with *facet_edges()***

```{r}
set_graph_style()

g <- ggraph(GAStech_graph, 
            layout = "nicely") + 
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department), 
                  size = 2)

g + facet_edges(~Weekday)
```

The code chunk below uses *theme()* to change the position of the legend.

```{r}
set_graph_style()

g <- ggraph(GAStech_graph, 
            layout = "nicely") + 
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department), 
                  size = 2) +
  theme(legend.position = 'bottom')
  
g + facet_edges(~Weekday)
```

### 4.5.2 **A framed facet graph**

```{r}
set_graph_style() 

g <- ggraph(GAStech_graph, 
            layout = "nicely") + 
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department), 
                  size = 2)
  
g + facet_edges(~Weekday) +
  th_foreground(foreground = "grey80",  
                border = TRUE) +
  theme(legend.position = 'bottom')
```

### **4.5.3 Working with *facet_nodes()***

```{r}
set_graph_style()

g <- ggraph(GAStech_graph, 
            layout = "nicely") + 
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department), 
                  size = 2)
  
g + facet_nodes(~Department)+
  th_foreground(foreground = "grey80",  
                border = TRUE) +
  theme(legend.position = 'bottom')
```

# **5 Network Metrics Analysis**

## **5.1 Computing centrality indices**

```{r}
g <- GAStech_graph %>%
  mutate(betweenness_centrality = centrality_betweenness()) %>%
  ggraph(layout = "fr") + 
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department,
            size=betweenness_centrality))
g + theme_graph()
```

-   *mutate()* of **dplyr** is used to perform the computation.

-   the algorithm used, on the other hand, is the *centrality_betweenness()* of **tidygraph**.

## **5.2 Visualising network metrics**

```{r}
g <- GAStech_graph %>%
  ggraph(layout = "fr") + 
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department, 
                      size = centrality_betweenness()))
g + theme_graph()
```

## **5.3 Visualising Community**

```{r}
g <- GAStech_graph %>%
  mutate(community = as.factor(group_edge_betweenness(weights = Weight, directed = TRUE))) %>%
  ggraph(layout = "fr") + 
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = community))  

g + theme_graph()
```

# **6 Building Interactive Network Graph with visNetwork**

-   [visNetwork()](http://datastorm-open.github.io/visNetwork/) is a R package for network visualization, using [vis.js](http://visjs.org/) javascript library.

-   *visNetwork()* function uses a nodes list and edges list to create an interactive graph.

    -   The nodes list must include an “id” column, and the edge list must have “from” and “to” columns.

    -   The function also plots the labels for the nodes, using the names of the actors from the “label” column in the node list.

## **6.1 Data preparation**

```{r}
GAStech_edges_aggregated <- GAStech_edges %>%
  left_join(GAStech_nodes, by = c("sourceLabel" = "label")) %>%
  rename(from = id) %>%
  left_join(GAStech_nodes, by = c("targetLabel" = "label")) %>%
  rename(to = id) %>%
  filter(MainSubject == "Work related") %>%
  group_by(from, to) %>%
    summarise(weight = n()) %>%
  filter(from!=to) %>%
  filter(weight > 1) %>%
  ungroup()
```

## **6.2 Plotting the first interactive network graph**

```{r}
visNetwork(GAStech_nodes, 
           GAStech_edges_aggregated)
```

## **6.3 Working with layout**

```{r}
visNetwork(GAStech_nodes,
           GAStech_edges_aggregated) %>%
  visIgraphLayout(layout = "layout_with_fr") 
```

## **6.4 Working with visual attributes - Nodes**

```{r}
GAStech_nodes <- GAStech_nodes %>%
  rename(group = Department) 
```

```{r}
visNetwork(GAStech_nodes,
           GAStech_edges_aggregated) %>%
  visIgraphLayout(layout = "layout_with_fr") %>%
  visLegend() %>%
  visLayout(randomSeed = 123)
```

## **6.5 Working with visual attributes - Edges**

```{r}
visNetwork(GAStech_nodes,
           GAStech_edges_aggregated) %>%
  visIgraphLayout(layout = "layout_with_fr") %>%
  visEdges(arrows = "to", 
           smooth = list(enabled = TRUE, 
                         type = "curvedCW")) %>%
  visLegend() %>%
  visLayout(randomSeed = 123)
```

## **6.6 Interactivity**

In the code chunk below, *visOptions()* is used to incorporate interactivity features in the data visualisation.

-   The argument *highlightNearest* highlights nearest when clicking a node.

-   The argument *nodesIdSelection* adds an id node selection creating an HTML select element.

```{r}
visNetwork(GAStech_nodes,
           GAStech_edges_aggregated) %>%
  visIgraphLayout(layout = "layout_with_fr") %>%
  visOptions(highlightNearest = TRUE,
             nodesIdSelection = TRUE) %>%
  visLegend() %>%
  visLayout(randomSeed = 123)
```
